First of all, I must thank my advisor, research supervisor, and thesis committee chair, Professor Samuel Guyer, for his unwavering
support. On a winter day, I walked into his office and asked if he had research projects; Sam was kind enough to offer me many
potential projects. After a few steers, now here we are. Although we faced many difficulties during the course of the three
semesters of research we did together, Sam has always been supportive as always.

I must also thank all my external collaborators, as Elephant Tracks II is the result of a collaborative effort and definitely not just
``my'' project. Especially, I would like to acknowledge the DaCapo Tooling Team at Google, including JC Beyler, Man Cao, Wessam Hassanein,
Kathryn McKinley, Ryan Rose, and Leandro Watanabe (in alphabetical order) as well as Steve Blackburn at the Australian National
University, for their useful insights and hard work.

I shall also thank Professor Kathleen Fisher, who, as the second reader on my thesis committee, the CS department chair, my programming
languages (COMP 105) professor, and a mentor of mine, was always helpful and caring. I need also to thank (now) Dr. Michael Shah, now
a lecturer at Northeastern University, who was once a marvelous teaching assistant who introduced me to computer science research;
without him, I might not have been here at all. Raoul Veroy, who is now the de facto maintainer of the original Elephant Tracks, also
provided me with a lot of assistance. Two fellow undergraduates, Jeremy Colebrook-Soucie and Matt Jones worked with me
on a different but closely related project, ``JumboViz'' (as I call it), and their work on that project also proved to be helpful
for the ET2 research I am doing.

There were a few other people who helped me on this project. I owe a lot to Luis Mastrangelo, who, as the author and maintainer of the
JNIF library, provided us generous help and even helped us debug our code. Ben Gamari and Kavon Farvadin helped with the part about
functional programming. Without them, it would have been really difficult for me to carry out my research and write my thesis.

Of course, there are many others important mentors and teachers whom I must thank. Without Professor Mark Sheldon's instruction
and mentorship, I wouldn't have gone this far in computer science, and especially not in programming languages. Other faculty
in the computer science department, such as Professor Norman Ramsey, Professor Noah Mendelsohn, Professor Ming Chow, and the
much beloved Professor Ben Hescott, have also supported me as an undergraduate at Tufts.

Many, many more people --- including, of course, my family and my friends --- supported me during the course of this project
and during my entire undergraduate career, but due to space limitations I could not possibly list all of them here, so I'd like to
apologize sincerely to all of those whom I have neglected to thank above. 
